---
title: "Assignment 2: Coding Basics"
author: "Cheney Gardner"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
seq1to100 <- seq(1, 100, 4)

mean(seq1to100); median(seq1to100)

mean(seq1to100) > median(seq1to100)

# I created a sequence with the function seq. 
# I then calculated summary statistics and  used a conditional statement.

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
names <- c("lucy", "bill", "luna", "jack")
scores <- c(80, 40, 90, 70)
pass <- c(TRUE, FALSE, TRUE, TRUE)

Student_Grades <- data.frame("Students"=names, "Scores"=scores, "Passed"=pass)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix can only be used if all of the inputs are the same type of data, but a data frame can store different types of data, like logical or numeric.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
passingtest <- function(x) {
  if(x > 50) {
    TRUE
  } 
  else if (x == 50) {
    TRUE
  }
  else {
    FALSE
  }
}

passingtest.2 <- function(x){
  ifelse(x>49, TRUE, FALSE)
  
}

passingtest(100)
passingtest(50)
passingtest(40)

print(passingtest(scores))

print(passingtest.2(scores))

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Because we're  looking at a vector of test scores i.e. the vector is length > 1, `ifelse` statements work. If we wanted to look at a single test score i.e. vector length = 1, I would use `if` and `else`.