---
title: "Assignment 3: Data Exploration"
author: "Cheney Gardner"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

```{r,  message = FALSE}
setwd("/Users/cheneygardner/Desktop/Env872/Environmental_Data_Analytics_2021")
getwd()
install.packages("ggplot2", repos="http://cran.us.r-project.org")
library(ggplot2)
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv")
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv")

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We might be interested in the ecotoxicology of neonicotinoids on insects because many insects are not detrimental to agriculture but in fact critical to its health, like bees. It is important to understand the effect of neonicotinoids on bees if we hope to preserve the contributions of polinators.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Leaf litter and woody debris in the forest provides important services, like releasing nutrients into the soil, helping sequester carbon and providing  hiding spaces for biodiversity. Litterfall and woody debris can also be used to estimate annual Aboveground Net Primary Productivity and aboveground biomass at plot, site, and continental scales, according to the NEON network guide.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Litter and fine wood debris is sampled at NEON sites that contain woody vegetation >2m tall
* Litter is defined as material that is dropped from the forest canopy and is collected in elevated 0.5m2 PVC traps.
* Fine wood debris is collected in ground traps.

## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)

```

> Answer: The neonics dataset contains 4623 observations of 30 variables.

6. Using the `summary` function on the "Effects" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(as.factor(Neonics$Effect))

```

> Answer: The most common effects studied are population, mortality and behavior. These effects might be of most interest because they are most easily observed and clear indicators of the effects of neonicotinoids on insects.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
summary(as.factor(Neonics$Species.Common.Name))

```

> Answer: The 6 most common species are: Honey Bee, Parasitic Wasp, Buff Tailed Bee, Bumble Bee, Carniolan Honey Bee, Bumble Bee and Italian Honeybee. The majority of these species are from the family Apidae and are known for their role in polination. Polinations is a critical to ensuring food production and security.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)

```

> Answer: The class of Conc.1..Author is a character vector because it refers to author defined criteria. 


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Neonics$Publication.Year))

geom_freqpoly

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Neonics$Publication.Year, color=Neonics$Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the field (natural) and labs. In this date range, the number of tests conducted in the field (natural) was highest around 2010 and the number of tests in the lab was highest around 2015. Since circa 2011, there have been many more tests in labs than in the field (natural).

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
endpoint.count <- Neonics$Endpoint
ggplot(Neonics, aes(x = endpoint.count)) +
  geom_bar()

```

> Answer: 
> LEOL - Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls

> NOEL - No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)
Litter$collectDate[1:10]
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)

unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)

```

> Answer: 14. Unique removes duplicate elements so that only "unique" elements are included, rather than including each individual element.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
litter.type<- Litter$functionalGroup
ggplot(Litter, aes(x = litter.type)) +
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
func.group <- Litter$functionalGroup
litter.mass <- Litter$dryMass
ggplot(Litter) +
  geom_boxplot(aes(x = func.group, y = litter.mass))

ggplot(Litter) +
  geom_violin(aes(x = func.group, y = litter.mass), 
              draw_quantiles = c(0.25, 0.5, 0.75))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is a more effective visualization for this data because we are looking at the biomass of different types of litter and do not need to visualize the density of data at different values. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles tend to have the highest biomass.