---
title: "Assignment 5: Data Visualization"
author: "Cheney Gardner"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r, tidy=TRUE}
#1 
setwd("/Users/cheneygardner/Desktop/Env872/Environmental_Data_Analytics_2021")
getwd()
library(tidyverse)
install.packages("cowplot",repos = "http://cran.us.r-project.org")
library(cowplot)
library(ggplot2)

peter.paul.chem <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
peter.paul.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
niwot.ridge.data <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 
class(peter.paul.chem$sampledate)
peter.paul.chem$sampledate <- as.Date(peter.paul.chem$sampledate, 
format = "%Y-%m-%d")
class(peter.paul.chem$sampledate)

class(peter.paul.nutrients$sampledate)
peter.paul.nutrients$sampledate <- as.Date(peter.paul.nutrients$sampledate, 
format = "%Y-%m-%d")
class(peter.paul.nutrients$sampledate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
new_theme <- theme_bw() + theme(panel.border = element_rect(color = "gray"), 
axis.text = element_text(size = 10), axis.title = element_text(size = 12))
theme_set(new_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
tp.ug.by.po4 <- ggplot(peter.paul.chem, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(size = 0.3) + 
  xlim(0, 50) +
  ylim(0, 100) +
  ylab("Phosphate") +
  xlab("Total Phospohrus") +
  ggtitle("Total phosphorus by phosphate") +
  geom_smooth(method = lm, se = FALSE, color = "black") +
  labs(color="Lake Name")
  print(tp.ug.by.po4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
##I'm unsure why the legend label isn't changing to "Lake Name" - 
#it changed when I originally wrote the code but now won't and I'm stumped

library(ggplot2)
temp.box.plot <-
  ggplot(peter.paul.chem, aes(x = as.factor(month), y = temperature_C, fill = lakename)) +
  geom_boxplot() +
  xlab("Month") +
  ylab("Temperature (C)") +
  ggtitle("Temp By Month") +
  labs(color="Lake Name")
print(temp.box.plot)

tp.box.plot <-
  ggplot(peter.paul.chem, aes(x = as.factor(month), y = tp_ug, fill = lakename)) +
  geom_boxplot() +
  xlab("Month") +
  ylab("Total Phosphorus") +
  labs(color = "Lake Name") +
  ggtitle("Total Phosphorus by Month")
print(tp.box.plot)

tn.box.plot <-
  ggplot(peter.paul.chem, aes(x = as.factor(month), y = tn_ug, fill = lakename)) +
  geom_boxplot() +
  xlab("Month") +
  ylab("TN") +
  ggtitle("TN By Month") +
  labs(color="Lake Name")
print(tn.box.plot)


library(cowplot) 
temp.tp.tn <- plot_grid(
temp.box.plot + theme(legend.position="none"), tp.box.plot + theme(legend.position="none"), tn.box.plot + theme(legend.position="none")
)

legend <- get_legend(temp.box.plot)
 
plot_grid(temp.tp.tn, legend, rel_widths = c(3, .4))
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
Concentrations of total phosphorus are highest for both lakes in late summer/ early fall (July, August, September). Peter lake has higher levels of phosphorus, month-by-month, than Paul Lake throughout the study period. TN concentration was relatively consistent throughout the year, but Peter Lake also has a higher concentration, month-by-month, in the study period. (Data for December and January unavailable.)

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
6.
install.packages("RColorBrewer",repos = "http://cran.us.r-project.org")
library(RColorBrewer)

litter.needles <-
  ggplot(subset(niwot.ridge.data, functionalGroup == "Needles"), 
         aes(x = collectDate, y = dryMass, color = nlcdClass)) + 
  geom_point() + 
  scale_shape_manual(values = c(15, 17)) +
  scale_color_brewer(palette = "Greens", direction = -1) +
   ylab("Dry Mass") +
  xlab("Collect Date") +
  labs(color="Dry Mass of Needle Litter by Date") +
  theme(legend.position = "right",
        legend.text = element_text(size = 12), legend.title = element_text(size = 12))
print(litter.needles)

7.
litter.needles.facet <-
  ggplot(subset(niwot.ridge.data, functionalGroup == "Needles"),
  aes(x = collectDate, y = dryMass)) + 
  ylab("Dry Mass") +
  xlab("Collect Date") +
  geom_point() + 
  facet_wrap(vars(nlcdClass))
  print(litter.needles.facet)
  
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
I prefer the plot showing dry mass of needle litter by date and separated by NLCD class using a color aesthetic because it provides a clear visual of major trends.This technique would likely not be as effective if the data was clustered or there were many overlying data points.
